Taking uncertainty in word embedding bias estimation seriously: a bayesian approach

### 2022-01-03 paperDraft/paperDraft4.tex
paper 4 started

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Taking uncertainty in word embedding bias estimation seriously: a bayesian approach},
  pdfauthor={Alicja Dobrzeniecka and Rafal Urbaniak},
  colorlinks=true,
  linkcolor={Maroon},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={blue},
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
%\documentclass{article}

@@

@@

Modern Natural Language Processing (NLP) models are used to complete
various tasks such as providing email filters, smart assistants, search
results, language translations, text analytics and so on. All of them
need as an input words represented by means of numbers which is
accomplished with word embeddings. It seems that in the learning process
these models can learn implicit biases that reflect harmful
stereotypical thinking. One of the sources of bias in NLP can be located
in the way the word embeddings are made. There is a considerable amount
of literature available on the topic of bias detection and mitigation in
NLP models.

One of the first measures in the discussion has been developed by
Bolukbasi, Chang, Zou, Saligrama, \& Kalai (2016). There, the gender
bias of a word \(w\) is understood as its projection on the gender
direction \(\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})\)
(the gender direction is the top principal compontent of ten gender pair
difference vectors). The underlying idea is that no bias is present if
non-explicitly gendered words are in equal distance to both elements in
all explicitly gender pairs. Given the (ideally) gender neutral words
\(N\) and the gender direction \(g\) the direct gender bias is defined
as the average distance of the words in \(N\) from \(g\) (\(c\) is a
parameter determining how strict we want to be): \begin{align}
\mathsf{directBias_c(N,g)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},g)\vert^c}{\vert N \vert }
\end{align}

The use of projections has been ciriticized for instance by Gonen \&
Goldberg (2019), who point out that while gender-direction might be an
indicator of bias, it is only one possible manifestation of it, and
reducing a projection of words might be insufficient. For instance,
``math'' and ``delicate'' might be in equal distance to both explicitly
gendered words while being closer to quite different stereotypical
attribute words. Further, the authors point out that most word pairs
preserve similarity under debiasing meant to minimize projection-based
bias.\footnote{Bolukbasi et al. (2016) use also another method which
  involves analogies and their evaluations by human users on Mechanical
  Turk. It is discussed and criticized in (Nissim, Noord, \& Goot,
  2020).}

To measure bias in word embeddings, Caliskan, Bryson, \& Narayanan
(2017) proposed the Word Embedding Association Test (WEAT). The idea is
that the measure of biases between two sets of target words, \(X\) and
\(Y\), (we call them protected words) should be quantified in terms of
the cosine similarity between the protected words and attribute words
coming from two sets of stereotype attribute words, \(A\) and \(B\)
(we'll call them attributes). For instance, \(X\) might be a set of male
names, \(Y\) a set of female names, \(A\) might contain stereotypically
male-related career words, and \(B\) stereotypically female-related
family words. WEAT is a modification of the Implicit Association Test
(IAT) (Nosek, Banaji, \& Greenwald, 2002) used in psychology and uses
almost the same word sets, allowing for a \emph{prima facie} sensible
comparison with bias in humans. If the person's attitude towards given
pair of concept is to be interpreted as neutral, there should be no
noticeable task completion time difference, and the final value from the
formula should be around 0. Let \(f\) be a similarity measure (usually,
cosine similarity). The association difference for a term \(t\) is:
\begin{align}
s(t,A,B) & = \frac{\sum_{a\in A}f(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}f(t,b)}{\vert B\vert}
\end{align} \noindent then, the association difference between \(A\) a
\(B\) is: \begin{align}
s(X,Y,A,B) & = \sum_{x\in X} s(x,A,B) -  \sum_{y\in Y} s(y,A,B)
\end{align} \noindent \(s(X,Y,A,B)\) is the statistic used in the
significance test, and the \(p\)-value obtained by bootstrapping: it is
the frequency of \(s(X_i,Y_i,A,B)>s(X,Y,A,B)\) for all equally sized
partitions \(X_i, Y_i\) of \(X\cup Y\). The effect size is computed by
normalizing the difference in means as follows: \begin{align}
bias(A,B) & = \frac{
\mu(\{s(x,A,B)\}_{x\in X}) -\mu(\{s(y,A,B)\}_{y\in Y}) 
}{
\sigma(\{s(w,A,B)\}_{w\in X\cup Y})
}
\end{align}

Caliskan et al. (2017) show that significant biases---thus measured---
similar to the ones discovered by IAT can be discovered in word
embeddings. Lauscher \& Glavas (2019) extended the methodology to a
multilingual and cross-lingual setting, arguing that using Euclidean
distance instead of similarity does not make much difference, while the
bias effects vary greatly across embedding models (interestingly, with
social media-text trained embeddings being less biased than those based
on Wikipedia).

A similar methodology is employed by Garg, Schiebinger, Jurafsky, \& Zou
(2018), who employ word embeddings trained on corpora from different
decades to study the shifts in various biases. For instance, to compute
the occupational embeddings bias for women the authors first compute the
average vector of vector embeddings of words that represent women
(e.g.~``she,'' ``female''), then calculate the Euclidean distance
between this mean vector and words for occupations. Then they take the
mean of these distances and subtract from it the analogously obtained
mean for the average vector of vector embeddings of words that represent
men. Formally they take the relative norm distance between \(X\) and
\(Y\) to be: \begin{align}
\textsf{relative norm distance} & = \sum_{v_m\in M} \vert \vert v_m - v_X\vert \vert_2 - \vert v_m - v_Y\vert \vert_2
\end{align} \noindent where the norm used is Euclidean, and \(v_X\) and
\(x_Y\) are average vectors for sets \(X\) and \(Y\) respectively.

Manzini, Lim, Tsvetkov, \& Black (2019) modify WEAT to a multi-class
setting, introducing Mean Average Cosine similarity as a measure of bias
(in fact, in the paper they report distances rather than similarities).
Let \(T = \{t_1, \dots, t_k\}\) be a class of protected word embeddings,
and let each \(A_j\in A\) be a set of attributes stereotypically
associated with a protected word). Then: \begin{align}
S(t_i, A_j) & = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cos}(t,a) \\
MAC(T,A) & = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} S(t_i,A_j)
\end{align} That is, for each protected word \$T and each attribute
class, they first take the mean for this protected word and all
attributes in a given attribute class, and then take the mean of thus
obtained means for all the protected words. The t-tests they employ are
run on average cosines used to calculate MAC.

@@

\hypertarget{some-methodological-problems}{%
\section{Some methodological
problems}\label{some-methodological-problems}}

Such approaches, however, are not unproblematic. One may consider the
selection of attributes for the bias research. For one thing, the word
lists are often fairly small, and the papers do not discuss the impact
of word list sizes on the uncertainty involved. We will argue that the
word list sizes are too small to yield strong conclusions, given the
variances involved.

We also observed that in the paper there is no word class distinction
and no control group. In the original paper, words from all three
religions were compared against all of the stereotypes, which means that
there was no distinction between cases in which the stereotype is
associated with a given religion, as opposed to the situation in which
it is associated with another one. Not all of the stereotypical words
have to be considered as harmful for all of the religions. One should
investigate the religions separately as some of them may have stronger
harmful associations that others. One should also include control groups
to have a way of comparing the stereotypical results with neutral or
human-like words. Later in the text we will explain in details reasons
for introducing control groups. In our analysis, we distinguish between
stereotypes associated with a given group, stereotypes associated with
different groups, and control groups: neutral words and stereotypes-free
human predicates.

The interpretation of the results is also a challenge. Assuming for a
moment that the value of multi-class cosine distance is correct, one may
question the interpretation. Manzini et al. (2019) summarize the
averages of cosine distance per group (gender, race, religion). For now
let us focus now on analyzing the values relating to religious biases.
Here is the relevant fragment of table:

\begin{longtable}[]{@{}ll@{}}
\toprule
Religion Debiasing & MAC \\
\midrule
\endhead
Biased & 0.859 \\
Hard Debiased & 0.934 \\
Soft Debiased (\(\lambda\) = 0.2) & 0.894 \\
\bottomrule
\end{longtable}

MAC stand for mean average cosine similarity, although in reality the
the table contains mean distances. What may attract attention is the
fact that the value of cosine distance in ``Biased'' category is already
quite high even before debiasing. High cosine distance indicates low
cosine similarity between values. One could think that average cosine
similarity equal to approximately 0.141 is not significant enough to
consider it as bias. However the authors aim to mitigate ``biases'' in
vectors with such great distance to make it even larger.
Methodologically the question is, on what basis is this small similarity
still considered as a proof of the presence of bias, and whether these
small changes are meaningful. This is in general the problem of scale
and the lack of universal intervals. In contrast, statistical intervals
will help us decide whether a given cosine similarity is high enough to
consider the words to be more similar than if we chose them at random.
We will use highest posterior density intervals, in line with Bayesian
methodology.

One may consider the issue of the curse of dimensionality. In our case,
the curse of dimensionality may take place when there is an increase in
the volume of data that results in adding extra dimensions to the
Euclidean space. As the number of features increases, it may be harder
and harder to obtain useful information from the data using the
available algorithms. Using cosine similarity in high dimensions in word
embeddings may be prone to the curse of dimensionality. According to
(\textbf{Venkat2018Curse?}) there are reasons to consider this
phenomenon when searching for word similarities in higher dimensions. An
experiment is conducted that aims at showing how the similarity values
and variation change as the number of dimensions increases. The
hypothesis made in the paper states that two things will happen as the
number of dimensions increase. First, the effort required to measure
cosine similarity will be greater, and two, the similarity between data
will blur out and have less variation. The authors generate random
points with increasing number of dimensions where each dimension of a
data point is given a value between 0 and 1. Then they pick one vector
at random from each dimension class and calculate the cosine similarity
between the chosen vector and the rest of the data. Then they check how
the variation of values changes as the number of dimensions increases.
It seems like the more dimensions there are, the smaller the variance
and therefore it is less obvious how to interpret the resulting cosine
similarities. Maybe the scale should be adjusted to the number of
dimensions and variance so that it still gives us sensible information
about data.

\todo{MAC: avoid looking at only one vector}

\todo{effect sizes}

\hypertarget{library-for-mac}{%
\section{Library for MAC}\label{library-for-mac}}

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-bolukbasi2016man}{}}%
Bolukbasi, T., Chang, K.-W., Zou, J., Saligrama, V., \& Kalai, A.
(2016). \emph{Man is to computer programmer as woman is to homemaker?
Debiasing word embeddings}. Retrieved from
\url{https://arxiv.org/abs/1607.06520}

\leavevmode\vadjust pre{\hypertarget{ref-Caliskan2017semanticsBiases}{}}%
Caliskan, A., Bryson, J. J., \& Narayanan, A. (2017). Semantics derived
automatically from language corpora contain human-like biases.
\emph{Science}, \emph{356}(6334), 183--186.
\url{https://doi.org/10.1126/science.aal4230}

\leavevmode\vadjust pre{\hypertarget{ref-Garg2018years}{}}%
Garg, N., Schiebinger, L., Jurafsky, D., \& Zou, J. (2018). Word
embeddings quantify 100 years of gender and ethnic stereotypes.
\emph{Proceedings of the National Academy of Sciences}, \emph{115}(16),
E3635--E3644. \url{https://doi.org/10.1073/pnas.1720347115}

\leavevmode\vadjust pre{\hypertarget{ref-Gonen2019lipstick}{}}%
Gonen, H., \& Goldberg, Y. (2019). Lipstick on a pig: {D}ebiasing
methods cover up systematic gender biases in word embeddings but do not
remove them. \emph{Proceedings of the 2019 Conference of the North
{A}merican Chapter of the Association for Computational Linguistics:
Human Language Technologies, Volume 1 (Long and Short Papers)},
609--614. Minneapolis, Minnesota: Association for Computational
Linguistics. \url{https://doi.org/10.18653/v1/N19-1061}

\leavevmode\vadjust pre{\hypertarget{ref-Lauscher2019multidimensional}{}}%
Lauscher, A., \& Glavas, G. (2019). Are we consistently biased?
Multidimensional analysis of biases in distributional word vectors.
\emph{CoRR}, \emph{abs/1904.11783}. Retrieved from
\url{http://arxiv.org/abs/1904.11783}

\leavevmode\vadjust pre{\hypertarget{ref-Manzini2019blackToCriminal}{}}%
Manzini, T., Lim, Y. C., Tsvetkov, Y., \& Black, A. W. (2019).
\emph{Black is to criminal as caucasian is to police: Detecting and
removing multiclass bias in word embeddings}. Retrieved from
\url{https://arxiv.org/abs/1904.04047}

>>>>

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Taking uncertainty in word embedding bias estimation seriously: a bayesian approach},
  pdfauthor={Alicja Dobrzeniecka and Rafal Urbaniak},
  colorlinks=true,
  linkcolor={Maroon},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={blue},
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
%\documentclass{article}

@@

\usepackage{sectsty}
\sectionfont{\fontsize{11}{11}\selectfont}

@@

Modern Natural Language Processing (NLP) models are used to complete
various tasks such as providing email filters, smart assistants, search
results, language translations, text analytics and so on. All of them
need as an input words represented by means of numbers which is
accomplished with word embeddings, in which particular lexical units are
repesented as vectors of real numbers. It has been suggested {[}1--6{]}
that in the learning process such models can learn implicit biases that
reflect harmful stereotypical thinking. A large chunk of the literature
on the topic focuses on the geometry of the learnt word embeddings; in
particular, unusually low (cosine) proximity of words belonging,
intuitively, to a stereotype, is taken as a sign that this stereotype
has been built into a given embedding. This methodology will be in the
focus of our paper.

One of the first measures in the discussion has been developed by
{[}1{]}. First, the gender direction \(gd\) in is obtained by taking the
differences of the vectors corresponding to ten different gendered pairs
(such as \(\overrightarrow{she} - \overrightarrow{he}\) or
\(\overrightarrow{girl} - \overrightarrow{boy}\)) and then identifying
their principal component (which is the vector obtained by projecting
the data points on their linear combination in a way that maximizes the
variance of the projections).\footnote{In the notebook associated with
  the paper, the authors simply use
  \(\overrightarrow{she} - \overrightarrow{he}\) as a gender direction,
  though.} The gender bias of a word \(w\) is understood as its
projection on the gender direction: \(\vec{w} \cdot gd\). Given the
supposedly gender neutral words \(N\)\footnote{We follow the methodology
  in assuming that there is a class of words that ideally should be
  identified as neutral, such as
  \emph{ballpark, solution, lecture, science, book} can be identified.
  We will have a bit to say about this assumption when we describe our
  dataset construction.} \todo{get back to this!} and the gender
direction \(gd\) the direct gender bias is defined as the average cosine
similarity of the words in \(N\) from \(gd\) (\(c\) is a parameter
determining how strict we want to be):

\footnotesize

\begin{align}
\mathsf{directBias_c(N,gd)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},gd)\vert^c}{\vert N \vert }
\end{align} \normalsize  The use of projections has been criticized for
instance by {[}4{]}, who point out that while the distance to the gender
direction might be an indicator of bias, it is only one possible
manifestation of it, and reducing the cosine distance to such a
projection might be insufficient. For instance, ``math'' and
``delicate'' might be in equal distance to a pair of opposed explicitly
gendered words, while being closer to quite different stereotypical
attribute words. Further, it is observed in {[}4{]} that most word pairs
preserve similarity under debiasing meant to minimize projection-based
bias.\footnote{In {[}1{]} another method which involves analogies and
  their evaluations by human users on Mechanical Turk is also used. We
  do not discuss this method in this paper, see its criticism in
  {[}7{]}.}

A measure of bias in word embeddings which does not employ gender
directions, the Word Embedding Association Test (WEAT), has been
proposed in {[}2{]}. The idea is that the bias between two sets of
target words, \(X\) and \(Y\) (we call them protected words), should be
quantified in terms of the cosine similarity between the protected words
and attribute words coming from two sets of stereotype attribute words,
\(A\) and \(B\) (we'll call them attributes). For instance, \(X\) might
be a set of male names, \(Y\) a set of female names, \(A\) might contain
stereotypically male-related, and \(B\) stereotypically female-related
career words. The association difference for a term \(t\) is:

\vspace{-2mm}

\footnotesize

\begin{align}
\mathsf{s}(t,A,B) & = \frac{\sum_{a\in A}\mathsf{cos}(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}\mathsf{cos}(t,b)}{\vert B\vert}
\end{align} \normalsize \noindent then, the association difference
between \(A\) a \(B\) is:

\vspace{-2mm}

\footnotesize

\begin{align}
\mathsf{s}(X,Y,A,B) & = \sum_{x\in X} \mathsf{s}(x,A,B) -  \sum_{y\in Y} \mathsf{s}(y,A,B)
\end{align} \normalsize

The effect size is computed by normalizing the difference in means as
follows:\footnote{ WEAT is a modification of the Implicit Association
  Test (IAT) {[}8{]} used in psychology, and it uses almost the same
  word sets, allowing for a \emph{prima facie} sensible comparison with
  bias in humans. {[}2{]} argue that significant biases---thus
  measured--- similar to the ones discovered by IAT can be discovered in
  word embeddings. {[}5{]} extended the methodology to a multilingual
  and cross-lingual setting, arguing that using Euclidean distance
  instead of similarity does not make much difference, while the bias
  effects vary greatly across embedding models (interestingly, with
  social media-text trained embeddings being less biased than those
  based on Wikipedia).}\(^{, }\)\footnote{ A similar methodology is
  employed in {[}3{]}. The authors employ word embeddings trained on
  corpora from different decades to study the shifts in various biases.
  For instance, to compute the occupational embeddings bias for women
  the authors first compute the average vector of vector embeddings of
  words that represent women (e.g.~\emph{she}, and \emph{female}), then
  calculate the Euclidean distance between this mean vector and words
  for occupations. Then they take the mean of these distances and
  subtract from it the analogously obtained mean for the average vector
  of vector embeddings of words that represent men. Formally they take
  the relative norm distance between \(X\) and \(Y\) to be:}

\vspace{-2mm}

\footnotesize

\begin{align}
\mathsf{weat}(A,B) & = \frac{
\mu(\{\mathsf{s}(x,A,B)\}_{x\in X}) -\mu(\{\mathsf{s}(y,A,B)\}_{y\in Y}) 
}{
\sigma(\{\mathsf{s}(w,A,B)\}_{w\in X\cup Y})
}
\end{align}

\normalsize

WEAT, however, has been developed to investigate biases corresponding to
a pair of supposedly opposing stereotypes, and so the question arises as
to how generalize the measure to contexts in which biases with respect
to more than two stereotypical groups are to be measured. Such a
generalization can be found in {[}6{]}. The authors introduce Mean
Average Cosine similarity as a measure of bias (strictly speaking, in
the paper they report cosine distances\footnote{By the cosine distance
  in the literature the authors mean 1-cosine similarity; note however
  that this terminology is slightly misleading, as mathematically it is
  not a distance measure, because it does not satisfy the triangle
  inequality, as generally \(dist(A,C) \not \leq dist(A,B)+ dist(B,C)\);
  we'll keep using this mainstream terminology though.} rather than
similarities). Let \(T = \{t_1, \dots, t_k\}\) be a class of protected
word embeddings, and let each \(A_j\in A\) be a set of attributes
stereotypically associated with a protected word). Then:

\vspace{-2mm }
\footnotesize

\begin{align}
\mathsf{s}(t_i, A_j) & = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cos}(t,a) \\
\mathsf{mac}(T,A) & = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} \mathsf{s}(t_i,A_j)
\end{align}

\normalsize

\noindent That is, for each protected word \(T\) and each attribute
class, they first take the mean for this protected word and all
attributes in a given attribute class, and then take the mean of thus
obtained means for all the protected words.

Having introduced the measures, first, we will introduce a selection of
general problems with this approach, and then we will move on to more
specific but important problems related to the statistical significance
of such measurement. We will focus on WEAT and MAC, as we want to put
issues with the use of projections aside. WEAT will be useful in our
criticism of the statistical methods involved, as it is simpler, so the
explanation and visualizations will be more transparent (and
\emph{mutatis mutandis} this criticism will apply to MAC as well), and
MAC will be useful in the development of our alternative method as its
range of applicability is the widest.

\hypertarget{non-statistical-problems-with-cosine-based-measures-of-bias}{%
\section{Non-statistical problems with cosine-based measures of
bias}\label{non-statistical-problems-with-cosine-based-measures-of-bias}}

One issue to consider is the selection of attributes for bias
measurement. The word lists used in the literature are often fairly
small (5-50)\todo{Check Manzani}. While the papers do employ statistical
tests to measure the uncertainty involved, we will later on argue that
these method are not proper for the goal at hand and show that a more
appropriate use of statistical methods leads to estimates of uncertainty
that are rather epistemologically pessimistic.
\todo{sensitivity to choice of words}

Let's think about MAC, using the case of religion-related
stereotypes.\todo{crossref to a list of words an explanation} In the
original paper, words from all three religions were compared against all
of the stereotypes. One reason this is problematic is that no
distinction between cases in which the stereotype is associated with a
given religion, as opposed to the situation in which it is associated
with another one, is made. This is problematic, as not all of the
stereotypical words have to be considered as harmful for all of the
religions. One should investigate the religions separately as some of
them may have stronger harmful associations that others.

The interpretation of the results is also a challenge. In {[}6{]} we can
find summaries of average cosine distances per group (such as gender,
race, or religion). For instance, for religion, here is the relevant
fragment of table:\todo{add p-values in the table}

\todo{kable the table}

\noindent (MAC stand for mean average cosine similarity, although in
reality the the table contains mean cosine distances). What may attract
attention is the fact that the value of cosine distance in ``Biased''
category is already quite high (i.e.~close to 1) even before debiasing.
High cosine distance indicates low cosine similarity between values. One
could think that average cosine similarity equal to approximately 0.141
is not significant enough to consider it as bias. However, the authors
still aim to mitigate such ``bias'' to make the distance even larger.
Methodologically the question is, on what basis is this small similarity
still considered as a proof of the presence of bias, and whether these
small changes are meaningful.

The underlying problem here is that in the paper there is no control
group. One should also include control groups to have a way of comparing
the results for a supposedly stereotyped group with the results for sets
of neutral or human-related neutral words. In our approach later on, we
distinguish between stereotypes associated with a given group,
stereotypes associated with different groups, and introduce control
groups: neutral words and stereotype-free human predicates.

\todo{bliskosc znaczeniowa a ko-okurencja}

\todo{comparison should be made to actual frequencies, to separate bias caused by cosine similarty to upstream bias}

\todo{unclear czy to sie przeklada na downstream effect}

\hypertarget{statistical-issues}{%
\section{Statistical issues}\label{statistical-issues}}

In contrast, statistical intervals will help us decide whether a given
cosine similarity is high enough to consider the words to be more
similar than if we chose them at random. We will use highest posterior
density intervals, in line with Bayesian methodology.

@@

\noindent \(s(X,Y,A,B)\) is the statistic used in the significance test,
and the \(p\)-value is obtained by bootstrapping: it is the frequency of
\(s(X_i,Y_i,A,B)>s(X,Y,A,B)\) for all equally sized partitions
\(X_i, Y_i\) of \(X\cup Y\). The effect size is computed by normalizing
the difference in means as follows:

\vspace{-2mm}

\footnotesize

\begin{align}
bias(A,B) & = \frac{
\mu(\{s(x,A,B)\}_{x\in X}) -\mu(\{s(y,A,B)\}_{y\in Y}) 
}{
\sigma(\{s(w,A,B)\}_{w\in X\cup Y})
}
\end{align}

\normalsize

The t-tests they employ are run on average cosines used to calculate
MAC.

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\vadjust pre{\hypertarget{ref-Bolukbasi2016man}{}}%
\CSLLeftMargin{{[}1{]} }
\CSLRightInline{Tolga Bolukbasi, Kai-Wei Chang, James Y. Zou, Venkatesh
Saligrama, and Adam Kalai. 2016. Man is to computer programmer as woman
is to homemaker? Debiasing word embeddings. \emph{CoRR} abs/1607.06520,
(2016). Retrieved from \url{http://arxiv.org/abs/1607.06520}}

\leavevmode\vadjust pre{\hypertarget{ref-Caliskan2017semanticsBiases}{}}%
\CSLLeftMargin{{[}2{]} }
\CSLRightInline{Aylin Caliskan, Joanna J. Bryson, and Arvind Narayanan.
2017. Semantics derived automatically from language corpora contain
human-like biases. \emph{Science} 356, 6334 (April 2017), 183--186.
DOI:https://doi.org/\href{https://doi.org/10.1126/science.aal4230}{10.1126/science.aal4230}}

\leavevmode\vadjust pre{\hypertarget{ref-Garg2018years}{}}%
\CSLLeftMargin{{[}3{]} }
\CSLRightInline{Nikhil Garg, Londa Schiebinger, Dan Jurafsky, and James
Zou. 2018. Word embeddings quantify 100 years of gender and ethnic
stereotypes. \emph{Proceedings of the National Academy of Sciences} 115,
16 (April 2018), E3635--E3644.
DOI:https://doi.org/\href{https://doi.org/10.1073/pnas.1720347115}{10.1073/pnas.1720347115}}

\leavevmode\vadjust pre{\hypertarget{ref-Gonen2019lipstick}{}}%
\CSLLeftMargin{{[}4{]} }
\CSLRightInline{Hila Gonen and Yoav Goldberg. 2019. Lipstick on a pig:
{D}ebiasing methods cover up systematic gender biases in word embeddings
but do not remove them. In \emph{Proceedings of the 2019 conference of
the north {A}merican chapter of the association for computational
linguistics: Human language technologies, volume 1 (long and short
papers)}, Association for Computational Linguistics, Minneapolis,
Minnesota, 609--614.
DOI:https://doi.org/\href{https://doi.org/10.18653/v1/N19-1061}{10.18653/v1/N19-1061}}

\leavevmode\vadjust pre{\hypertarget{ref-Lauscher2019multidimensional}{}}%
\CSLLeftMargin{{[}5{]} }
\CSLRightInline{Anne Lauscher and Goran Glavas. 2019. Are we
consistently biased? Multidimensional analysis of biases in
distributional word vectors. \emph{CoRR} abs/1904.11783, (2019).
Retrieved from \url{http://arxiv.org/abs/1904.11783}}

\leavevmode\vadjust pre{\hypertarget{ref-Manzini2019blackToCriminal}{}}%
\CSLLeftMargin{{[}6{]} }
\CSLRightInline{Thomas Manzini, Yao Chong Lim, Yulia Tsvetkov, and Alan
W Black. 2019. Black is to criminal as caucasian is to police: Detecting
and removing multiclass bias in word embeddings. Retrieved from
\url{https://arxiv.org/abs/1904.04047}}

\leavevmode\vadjust pre{\hypertarget{ref-Nissim2020fair}{}}%
\CSLLeftMargin{{[}7{]} }
\CSLRightInline{Malvina Nissim, Rik van Noord, and Rob van der Goot.
2020. Fair is better than sensational: Man is to doctor as woman is to
doctor. \emph{Computational Linguistics} 46, 2 (June 2020), 487--497.
DOI:https://doi.org/\href{https://doi.org/10.1162/coli_a_00379}{10.1162/coli\_a\_00379}}

\leavevmode\vadjust pre{\hypertarget{ref-Nosek2002harvesting}{}}%
\CSLLeftMargin{{[}8{]} }
\CSLRightInline{Brian A. Nosek, Mahzarin R. Banaji, and Anthony G.
Greenwald. 2002. Harvesting implicit group attitudes and beliefs from a
demonstration web site. \emph{Group Dynamics: Theory, Research, and
Practice} 6, 1 (2002), 101--115.
DOI:https://doi.org/\href{https://doi.org/10.1037/1089-2699.6.1.101}{10.1037/1089-2699.6.1.101}}
